### 2017-04-10 paper/sexSelectionSpacing-ver1.tex
Editing hook in introduction

Understanding and measuring son preference has been an important research 
area for decades and---with increases in many countries' male-to-female ratios 
at birth---is as salient as ever.
India, for example, has seen

Initial research relied on how birth spacing and the likelihood of stopping 
childbearing differed across births of sons versus daughters to establish 
son preference.
With the introduction of prenatal sex determination the sex of children
born was

%
\footnote{
See \citet{das_gupta97}, \citet{Sudha1999},
\citet{Arnold2002}, \cite{retherford03b} and \citet{jha06}.
India is not alone; both China and South Korea saw
significant changes in the sex ratio at birth \citep{Yi1993,park95}.
}

Our understanding of the use of sex selection is, however, constrained by 
lack of information;
there are no official data on sex selection, and the few surveys that ask about use of 
sex selection show signs of serious under-reporting \citep{goodkind96}.

With no direct information, researchers have relied on a simple method for establishing 
factors that affect sex selection: use the sex of children born as the 
dependent variable, and estimate the effects of variables on the probability of having a son.%
\footnote{
In the absence of any interventions, the probability of having a son
is approximately 0.512, which is independent of genetic factors \citep{ben-porath76b,jacobsen99}.
With fetus sex random, a statistically significant variable indicates an association with 
use of sex-selective abortions.
Examples of studies that have used this approach are \cite{retherford03b},
\cite{jha06}, and \cite{abrevaya09}. 
}
Based on this simple method, we know that families with no sons are more likely to
use sex selection the higher the parity;
that use of sex selection increases with socioeconomic status, especially education;
and that sex selection is more widespread in cities than in rural areas 
\citep{retherford03b,jha06,abrevaya09}.%
\footnote{
There is substantial disagreement on whether sex-selective abortion is used for the 
first birth \citep{retherford03b,jha06}.
}

@@

In the absence of sex-selective abortions, son preference often leads to a shorter duration 
until the next birth if the previous birth was a daughter  
\citep[see, for example,][]{Das1987,Rahman1993,Pong1994,Haughton1996,Arnold1997}.
The resulting shorter spacing is thought to be associated with worse health outcomes for
the girls \citep{arnold98,Whitworth2002,Rutstein2005,Conde-Agudelo2006}.

@@

Recent research suggests that son preference in India, when measured as ideally having 
more boys than girls, is decreasing over time and with higher education \citep{bhat03,pande07}.%
\footnote{
This measure of son preference is commonly used in the literature. 
See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
In NFHS, and Demographic and Health Surveys in general, son preference is measured as part 
of the general fertility questions.
The first question is how many children a woman would have if she could choose exactly how 
many.
This if followed by the sex preference question:
``How many of these children would you like to be boys, how many would you like to be girls 
and for how many would the sex not matter?''
}

@@

In the absence of prenatal sex determination, parents' only choice if they 
wanted a son was to try for another pregnancy as soon as possible after
the birth of a girl. 
This led to shorter spacing after the birth of a girl and what
is known as differential stopping behavior, where parents were 
more likely to cease childbearing after the birth of a son than
after a daughter [TK add references here]. 
As discussed above, the shorter spacing after the birth of a girl is 
associated with worse health outcomes for both mother and children.
Because of the potential health effects, most of the literature on spacing 
for countries like India naturally focus on whether spacing was too short
and used the shorter birth spacing after the birth of a girl than after the
birth of a boy as a measure of the degree of son preference.

>>>>

Understanding and measuring son preference has been an important research 
area for decades and is as salient as ever.
India, for example, has seen dramatic increases in the males-to-females ratio 
at birth over the last three decades as access to prenatal sex determination 
expanded.%
\footnote{
See \citet{das_gupta97}, \citet{Sudha1999},
\citet{Arnold2002}, \cite{retherford03b} and \citet{jha06}.
India is not alone; both China and South Korea saw
significant changes in the sex ratio at birth \citep{Yi1993,park95}.
}
Recent research suggests, however, that son preference in India, when measured 
as ideally having more boys than girls, is decreasing over time and with 
higher education \citep{bhat03,pande07}.%
\footnote{
This measure of son preference is commonly used in the literature. 
See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
}

In the absence of prenatal sex determination, parents' only choice if they 
want a son is to try for another pregnancy as soon as possible after
the birth of a girl. 
Hence, in the absence of sex-selective abortions, son preference  
leads to a shorter duration until the next birth if the previous 
birth was a daughter  
\citep[see, for example,][]{Das1987,Rahman1993,Pong1994,Haughton1996,Arnold1997}.
The resulting shorter spacing is thought to be associated with worse 
health outcomes for the girls 
\citep{arnold98,Whitworth2002,Rutstein2005,Conde-Agudelo2006}.
Parents are also more likely to cease childbearing after the birth of 
a son than after a daughter \citep{ben-porath76b,Das1987,Arnold1997,clark00}.

With the introduction of sex-selective abortions, parents can now abort
a fetus of unwanted sex and the use of sex selection can therefore 
directly provide us a picture of the degree of son preference.
Our understanding of the use of sex selection is, however, constrained by 
lack of information;
there are no official data on sex selection, and the few surveys that ask 
about use of sex selection show signs of serious under-reporting \citep{goodkind96}.
With no direct information, researchers have relied on a simple method for 
establishing factors that affect sex selection: use the sex of children born as 
the dependent variable, and estimate the effects of variables on the probability 
of having a son.%
\footnote{
In the absence of any interventions, the probability of having a son
is approximately 0.512, which is independent of genetic factors \citep{ben-porath76b,jacobsen99}.
With fetus sex random, a statistically significant variable indicates an association with 
use of sex-selective abortions.
Examples of studies that have used this approach are \cite{retherford03b},
\cite{jha06}, and \cite{abrevaya09}. 
}
Based on this simple method, we know that families with no sons are more likely to
use sex selection the higher the parity;
that use of sex selection increases with socioeconomic status, especially education;
and that sex selection is more widespread in cities than in rural areas 
\citep{retherford03b,jha06,abrevaya09}.%
\footnote{
There is substantial disagreement on whether sex-selective abortion is used for the 
first birth \citep{retherford03b,jha06}.
}

Because of the potential health effects, most of the literature on spacing 
for countries like India naturally focus on whether spacing was too short
and used the shorter birth spacing after the birth of a girl than after the
birth of a boy as a measure of the degree of son preference.

@@

@@

@@
